---
title: "Visualizing Student Debt:"
subtitle: "Student Loans, State Spending, and Tuition" 
author: 
  - Natalie Meltzer
date: '`r Sys.Date()`'
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: true
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.width=9, fig.height=3.5, fig.retina=3,
  out.width = "100%",
  cache = FALSE,
  echo = TRUE,
  message = FALSE, 
  warning = FALSE,
  hiline = TRUE
)

```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
library(sysfonts)
style_mono_light(base_color = "#23395b",
  header_font_google = google_font("Mulish", "600"),
  text_font_google   = google_font("Mulish", "300", "300i"),
  code_font_google   = google_font("Mulish"))
style_xaringan(colors = c(
  red = "#f34213",
  purple = "#3e2f5b",
  orange = "#ff8811",
  green = "#136f63",
  white = "#FFFFFF"))
  
```

```{r data, include=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
library(janitor)
library(patchwork)
library(usdata)
library(maps)
library(ggplot2)
  

# read in data with multiple sheets 
# code from https://stackoverflow.com/questions/49359587/import-excel-workbook-with-multiple-sheets

path <- "../data/household_debt.xlsx"
sheetnames <- excel_sheets(path)
debtlevels <- lapply(excel_sheets(path), read_excel, path = path)
names(debtlevels) <- sheetnames

studentdebt <- lapply(excel_sheets(path), read_excel, path = path)
names(debtlevels) <- sheetnames

path2 <- "../data/state_spending.xls"
sheetnames <- excel_sheets(path2)
statespending <- lapply(excel_sheets(path2), read_excel, path = path2)
names(statespending) <- sheetnames


# read in other data 


studentdebt <- read_excel("../data/totaldebtbalance.xlsx") %>%
  select(c("Year", "Student Loan"))%>%
  clean_names() %>%
  filter(grepl("Q4", year))%>%
  mutate("year" = str_sub(year, 1,2)) %>%
  mutate("year" = str_glue("20{year}"))%>%
    mutate("year" = as.numeric(year)) %>%
   mutate("studentloan_adj" = case_when (
    year == 2004 ~ student_loan * 1.35,
     year == 2005 ~ student_loan * 1.31,
     year == 2006 ~ student_loan * 1.27,
     year == 2007 ~ student_loan * 1.22,
     year == 2008 ~ student_loan * 1.22,
     year == 2009 ~ student_loan * 1.19,
     year == 2010 ~ student_loan * 1.17,
     year == 2011 ~ student_loan * 1.14,
     year == 2012 ~ student_loan * 1.12,
     year == 2013 ~ student_loan * 1.1,
     year == 2014 ~ student_loan * 1.09,
     year == 2015 ~ student_loan * 1.09,
     year == 2016 ~ student_loan * 1.06,
     year == 2017 ~ student_loan * 1.04,
     year == 2018 ~ student_loan * 1.02,
     year == 2019 ~ student_loan * 1))%>%
  filter(!is.na(studentloan_adj))%>%
  mutate("studentloan_adj" = 1000*studentloan_adj)

  

educationcosts <- read_excel("../data/education_costs.xls") %>% 
  clean_names() %>%
 mutate("year" = str_sub(year, 1,4))

studentdebt <- read_excel("../data/totaldebtbalance.xlsx") %>%
  select(c("Year", "Student Loan"))%>%
  clean_names() %>%
  filter(grepl("Q4", year))%>%
  mutate("year" = str_sub(year, 1,2)) %>%
  mutate("year" = str_glue("20{year}"))%>%
    mutate("year" = as.numeric(year)) %>%
   mutate("studentloan_adj" = case_when (
    year == 2004 ~ student_loan * 1.35,
     year == 2005 ~ student_loan * 1.31,
     year == 2006 ~ student_loan * 1.27,
     year == 2007 ~ student_loan * 1.22,
     year == 2008 ~ student_loan * 1.22,
     year == 2009 ~ student_loan * 1.19,
     year == 2010 ~ student_loan * 1.17,
     year == 2011 ~ student_loan * 1.14,
     year == 2012 ~ student_loan * 1.12,
     year == 2013 ~ student_loan * 1.1,
     year == 2014 ~ student_loan * 1.09,
     year == 2015 ~ student_loan * 1.09,
     year == 2016 ~ student_loan * 1.06,
     year == 2017 ~ student_loan * 1.04,
     year == 2018 ~ student_loan * 1.02,
     year == 2019 ~ student_loan * 1))%>%
  filter(!is.na(studentloan_adj))%>%
  mutate("studentloan_adj" = 1000*studentloan_adj)


# clean data
## convert lists to tibble, drop unnecessary columns, filter unnecessary rows



debtlevels <- enframe(debtlevels) %>% 
  select(name, value) %>%
  unnest_wider(value) %>%
  select(!ends_with(c("_2003", "_2020","_2021"))) %>%
  rename("debt_type" = name)



debtlevels  <- debtlevels %>% unnest(cols = c(state:Q4_2019))%>%
  filter(debt_type != "total") %>%
  filter(state != "allUS") %>%
   pivot_longer(!debt_type:state, names_to = "year", values_to = "amount") %>%
  pivot_wider(names_from = debt_type, values_from = amount)%>%
   mutate(year = str_remove(year, "Q4_")) %>%
  select(c("state", "year", "studentloan", "studentloan_delinq"))
  

statespending <- enframe(statespending, name = "year") %>%
  select(year, value) %>%
  unnest_longer(value) %>%
  unnest(cols = "value") %>% 
  clean_names() %>%
  select(c("state", "year", "higher_education"))

## merge datasets 

debtspending <- debtlevels %>%
  mutate(state = abbr2state(state)) %>%
  right_join(statespending, by = c("state", "year")) 


## adjust for inflation - multipliers from BLS

debtspending <- debtspending %>% 
  mutate("studentloan_adj" = case_when (
    year == 2004 ~ studentloan * 1.35,
     year == 2005 ~ studentloan * 1.31,
     year == 2006 ~ studentloan * 1.27,
     year == 2007 ~ studentloan * 1.22,
     year == 2008 ~ studentloan * 1.22,
     year == 2009 ~ studentloan * 1.19,
     year == 2010 ~ studentloan * 1.17,
     year == 2011 ~ studentloan * 1.14,
     year == 2012 ~ studentloan * 1.12,
     year == 2013 ~ studentloan * 1.1,
     year == 2014 ~ studentloan * 1.09,
     year == 2015 ~ studentloan * 1.09,
     year == 2016 ~ studentloan * 1.06,
     year == 2017 ~ studentloan * 1.04,
     year == 2018 ~ studentloan * 1.02,
     year == 2019 ~ studentloan * 1)) %>%
  mutate("highered_adj" = case_when (
    year == 2004 ~ higher_education * 1.35,
     year == 2005 ~ higher_education * 1.31,
     year == 2006 ~ higher_education * 1.27,
     year == 2007 ~ higher_education * 1.22,
     year == 2008 ~ higher_education * 1.22,
     year == 2009 ~ higher_education * 1.19,
     year == 2010 ~ higher_education * 1.17,
     year == 2011 ~ higher_education * 1.14,
     year == 2012 ~ higher_education * 1.12,
     year == 2013 ~ higher_education * 1.1,
     year == 2014 ~ higher_education * 1.09,
     year == 2015 ~ higher_education * 1.09,
     year == 2016 ~ higher_education * 1.06,
     year == 2017 ~ higher_education * 1.04,
     year == 2018 ~ higher_education * 1.02,
     year == 2019 ~ higher_education * 1)) %>%
  select(-c("studentloan", "higher_education"))
   

debtspending <- debtspending %>%
  group_by(state) %>%
  mutate(loan_pct_chg = 100 *((studentloan_adj-lag(studentloan_adj))/lag(studentloan_adj))) %>%
   mutate(spend_pct_chg = 100 *((highered_adj-lag(highered_adj))/lag(highered_adj)))

statemap <- map_data("state") %>%
  mutate(state = str_to_title(region, locale = "en"))
  
debtspending <- debtspending %>%
  left_join(statemap, by = "state") %>%
  filter(!is.na(long))
```


# **How many of you have student loans?**
--

### **.red[42]** million Americans — about **.red[1]** in every **.red[8]** — have student loans

.footnote[Adam Looney, David Wessel, and Kadija Yilla, Brookings Institute]
--

### **.red[75%]** of student loan borrowers took out loans to go to two- or four-year colleges — representing about 50% of outstanding debt

--

### **.red[25%]** of student loan borrowers took out loans for grad school — representing the other 50% of outstanding debt



---

# It has not always been this way


```{r echo = FALSE}
ggplot(studentdebt, aes(x = year, 
            y = studentloan_adj))+
  geom_line()+
  scale_y_continuous(labels = scales::dollar_format())+
  theme_minimal()+
    theme_xaringan(background_color = "#FFFFFF") +
  theme(plot.title.position = "plot",
        plot.title = element_text(size=20,
                                  face="bold"),
        plot.subtitle = element_text(size=15),
        plot.caption = element_text(size=10))+
  labs(title = "Total student loan burden",
       subtitle = "Dollars outstanding (in billions, $2019)",
       caption = "Data: Federal Reserve Bank of New York, Quarterly Report on Household Debt and Credit",
       x = "",
       y = "")

```

---

# It has not always been this way


```{r echo = FALSE}

debtspending %>%
  group_by(year, state) %>%
  summarise(avg_loan = mean(studentloan_adj), na.rm = TRUE) %>%
  ungroup() %>%
  group_by(year) %>%
  summarise(avg_loan_year = mean(avg_loan), na.rm = TRUE)%>%
  ungroup()%>%
  mutate(year = as.numeric(year))%>%
ggplot(aes(x = year, 
            y = avg_loan_year))+
  geom_line()+
  scale_y_continuous(labels = scales::dollar_format())+
  theme_minimal()+
    theme_xaringan(background_color = "#FFFFFF") +
  theme(plot.title.position = "plot",
        plot.title = element_text(size=20,
                                  face="bold"),
        plot.subtitle = element_text(size=15),
        plot.caption = element_text(size=10))+
  labs(title = "Average student loan per capita",
       subtitle = "($2019)",
       caption = "Data: Federal Reserve Bank of New York, Quarterly Report on Household Debt and Credit",
       x = "",
       y = "")

```

---

# Is the increase in student debt a problem? 


```{r, echo = FALSE}
debtspending %>%
  group_by(year, state) %>%
  summarise(delinq = mean(studentloan_delinq), na.rm = TRUE) %>%
  ungroup() %>%
  group_by(year) %>%
  summarise(avg_delinq = mean(delinq), na.rm = TRUE)%>%
  ungroup()%>%
  mutate(year = as.numeric(year))%>%
  mutate(avg_delinq = avg_delinq/100)%>%
ggplot(aes(x = year, 
            y = avg_delinq))+
  geom_line()+
  scale_y_continuous(labels = scales::percent_format())+
  theme_minimal()+
    theme_xaringan(background_color = "#FFFFFF") +
  theme(plot.title.position = "plot",
        plot.title = element_text(size=20,
                                  face="bold"),
        plot.subtitle = element_text(size=15),
        plot.caption = element_text(size=10))+
  labs(title = "Average student loan delinquency rate",
       caption = "Data: Federal Reserve Bank of New York, Quarterly Report on Household Debt and Credit",
       x = "",
       y = "")

```


---

# Tuition and fees have increased significantly


```{r echo = FALSE}

educationcosts %>%
  filter(institution_type != "Public and Private") %>%
  ggplot()+
  geom_line(aes(x = year, 
               y = all_institutions, 
               color = institution_type,
               group = institution_type))+
  scale_y_continuous(labels = scales::dollar_format())+
  scale_color_discrete(name = "Institution Type")+
  theme_minimal()+
  theme_xaringan(background_color = "#FFFFFF")+
  theme(legend.position = "right",
    plot.title.position = "plot",
        plot.title = element_text(size=20,
                                  face="bold"),
        plot.subtitle = element_text(size=15),
        plot.caption = element_text(size=10),
    legend.title = element_text(size=15),
    legend.text = element_text(size=15))+
  scale_x_discrete(breaks=c("2004","2006","2008","2010","2012","2014","2016","2018"))+
  labs(title = "Annual Tuition and Fees",
       subtitle = "Private and Public 2- and 4- Year Colleges ($2019)",
       caption = "Data: U.S. Department of Education, Integrated Postsecondary Education Data System, Institutional Characteristics component",
       x = "",
       y = "")
  

```


---

# State spending on higher education has decreased


```{r, echo = FALSE}
debtspending %>%
  group_by(year, state) %>%
  summarise(avg_spend = mean(highered_adj), na.rm = TRUE) %>%
  ungroup() %>%
  group_by(year) %>%
  summarise(avg_percap_spend = mean(avg_spend), na.rm = TRUE)%>%
  ungroup()%>%
  mutate(year = as.numeric(year))%>%
ggplot(aes(x = year, 
            y = avg_percap_spend))+
  geom_line()+
  scale_y_continuous(labels = scales::dollar_format())+
  theme_minimal()+
    theme_xaringan(background_color = "#FFFFFF") +
  theme(plot.title.position = "plot",
        plot.title = element_text(size=20,
                                  face="bold"),
        plot.subtitle = element_text(size=15),
        plot.caption = element_text(size=10))+
  labs(title = "Average State Spending on Higher Education Per Capita",
       subtitle = "$2019",
       caption = "Data: U.S. Census Bureau, Annual Survey of State and Local Government Finances\n Compiled by the Urban-Brookings Tax Policy Center",
       x = "",
       y = "")
  
```
---

# Breaking it down by state


```{r, echo = FALSE}
debtspending %>%
  filter(year == 2004) %>%
  ggplot(aes(x = long, y = lat, group = group))+
  geom_polygon(aes(fill = studentloan_adj), color = "black")+
   scale_fill_gradient(low = "white", 
                        high = "red",
                        name = "Avg. student loans per capita ($2019)",
                        label = scales::dollar_format(),
                       limits = c(1000,8000)) + 
  theme(legend.position = "right",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        rect = element_blank(),
        plot.title.position = "plot",
        plot.title = element_text(size=20,
                                  face="bold"),
        plot.caption = element_text(size=10))+
  labs(title = "Average Student Loans Per Capita, 2004",
       caption = "Data: Federal Reserve Bank of New York, Quarterly Report on Household Debt and Credit")
 

```

---

# Breaking it down by state

```{r, echo = FALSE}
debtspending %>%
  filter(year == 2019) %>%
  ggplot(aes(x = long, y = lat, group = group))+
  geom_polygon(aes(fill = studentloan_adj), color = "black")+
  scale_fill_gradient(low = "white", 
                        high = "red",
                        name = "Avg. student loans per capita ($2019)",
                        label = scales::dollar_format(),
                      limits = c(1000,8000)) + 
  theme(legend.position = "right",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        rect = element_blank(),
        plot.title.position = "plot",
        plot.title = element_text(size=20,
                                  face="bold"),
        plot.caption = element_text(size=10))+
  labs(title = "Average Student Loans Per Capita, 2019",
       caption = "Data: Federal Reserve Bank of New York, Quarterly Report on Household Debt and Credit")

```

---

# Breaking it down by state

```{r, echo = FALSE}
debtspending %>%
  filter(year == 2004) %>%
  ggplot(aes(x = long, y = lat, group = group))+
  geom_polygon(aes(fill = studentloan_delinq), color = "black")+
   scale_fill_gradient(low = "white", 
                        high = "red",
                        name = "% Delinquent",
                        limits = c(1,20)) + 
  theme(legend.position = "right",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        rect = element_blank(),
        plot.title.position = "plot",
        plot.title = element_text(size=20,
                                  face="bold"),
        plot.caption = element_text(size=10))+
  labs(title = "Student Loan Delinquency Rate, 2004",
       caption = "Data: Federal Reserve Bank of New York, Quarterly Report on Household Debt and Credit")
 
```

---

# Breaking it down by state

```{r, echo = FALSE}
debtspending %>%
  filter(year == 2019) %>%
  ggplot(aes(x = long, y = lat, group = group))+
  geom_polygon(aes(fill = studentloan_delinq), color = "black")+
   scale_fill_gradient(low = "white", 
                        high = "red",
                        name = "% Delinquent",
                      limits = c(1,20) ) + 
  theme(legend.position = "right",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        rect = element_blank(),
        plot.title.position = "plot",
        plot.title = element_text(size=20,
                                  face="bold"),
        plot.caption = element_text(size=10))+
  labs(title = "Student Loan Delinquency Rate, 2019",
       caption = "Data: Federal Reserve Bank of New York, Quarterly Report on Household Debt and Credit")
 
```

---

# Digging deeper in a dashboard 

[**My beautiful dashboard**](https://nameltz.shinyapps.io/dashboard/)<br>

---
class: center, middle

# Thanks!

Slides created via the R packages:

[**xaringan**](https://github.com/yihui/xaringan)<br>
[gadenbuie/xaringanthemer](https://github.com/gadenbuie/xaringanthemer).


